Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2017-01-26 manuscript.tex
latest

In the classic lottery model \citep{chesson_1981}, unoccupied territories are assumed to be saturated with propagules from every genotype i.e. $l_i\gg 1$ where $l_i=m_i/U$ is the mean propagule density. From the law of large numbers, the composition of propagules in each territory will then not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$ ($G$ is the number of genotypes present), and so the probability that genotype $i$ wins any particular unoccupied territory is approximately $c_i l_i/\sum_j c_j l_j$. Thus, genotype $i$ is expected to win $c_i l_i/\sum_j c_j l_j$ of the $U$ available territories, and deviations from this expected outcome are small (since $T$ is large by assumption). Consequently, the increase in $n_i$ due to territorial acquisition in the classic lottery model is given by
\begin{equation}
\Delta_+ n_i(t)=\frac{c_i l_i}{\sum_j c_j l_j}U(t)=b_i n_i\frac{1}{L}\frac{c_i}{\overline{c}}, \label{eq:lottery}
\end{equation}
where $\overline{c}=\sum_j c_j m_j/M$ is the mean propagule competitive ability for a randomly selected propagule, $L=M/U$ is the total propagule density and $M=\sum_j m_j$ is the total number of propagules. Eq. \eqref{eq:lottery} is nonsensical at low densities, specifically if $l_i$ is sufficiently small: genotype $i$ can win at most $m_i$ territories, yet Eq. \eqref{eq:lottery} demands a fraction $c_i l_i/\sum_j c_j l_j$ of the unoccupied territories $U$, no matter how large $U$ is.

@@

W

%Thus, the numbers of territorial victories $\Delta_+ n_1,\Delta_+ n_2,\ldots$ follow a multinomial distribution with $U$ trials and event probabilities $\frac{c_1 l_1}{\sum_j c_j l_j},\frac{c_2 l_2}{\sum_j c_j l_j},\ldots$.

>>>>

In the classic lottery model \citep{chesson_1981}, unoccupied territories are assumed to be saturated with propagules from every genotype i.e. $l_i\gg 1$ where $l_i=m_i/U$ is the mean propagule density. From the law of large numbers, the composition of propagules in each territory will then not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$ ($G$ is the number of genotypes present), and so the probability that genotype $i$ wins any particular unoccupied territory is approximately $c_i l_i/\sum_j c_j l_j$. Let $\Delta_+ n_i$ denote the number of territories won by genotype $i$ in the current iteration. Then $\Delta_+ n_1,\Delta_+ n_2,\ldots,\Delta_+ n_G$ follow a multinomial distribution with $U$ trials and success probabilities $\frac{c_1 l_1}{\sum_j c_j l_j},\frac{c_2 l_2}{\sum_j c_j l_j},\ldots,\frac{c_G l_G}{\sum_j c_j l_j}$, respectively. Genotype $i$ is expected to win $c_i l_i/\sum_j c_j l_j$ of the $U$ available territories, and deviations from this expected outcome are small (since $T$ is large by assumption), giving 
\begin{equation}
\Delta_+ n_i(t)=\frac{c_i l_i}{\sum_j c_j l_j}U(t)=b_i n_i\frac{1}{L}\frac{c_i}{\overline{c}}, \label{eq:lottery}
\end{equation}
where $\overline{c}=\sum_j c_j m_j/M$ is the mean propagule competitive ability for a randomly selected propagule, $L=M/U$ is the total propagule density and $M=\sum_j m_j$ is the total number of propagules. Eq. \eqref{eq:lottery} is nonsensical at low densities, specifically if $l_i$ is sufficiently small: genotype $i$ can win at most $m_i$ territories, yet Eq. \eqref{eq:lottery} demands a fraction $c_i l_i/\sum_j c_j l_j$ of the unoccupied territories $U$, no matter how large $U$ is.

The classic lottery model is closely related to the Wright-Fisher model of genetic drift. In the Wright-Fisher model, genotype abundances in each generation follow a multinomial distribution with success probabilities $w_i n_i/\sum_j w_j n_j$, where $w$ is relative fitness and the $n_i$ are  genotype abundances in the preceding generation. Population size $N$ remains constant. This is mathematically equivalent to the classic lottery model with non-overlapping generations ($d_i=1$ for all $i$) and $w_i=b_i c_i$. The classic lottery model is more general in that it allows for overlapping generations, which crucially introduces ``storage'' effects \citep{svardal_2015}. Additionally, the relative fitnesses $w_i=b_i c_i$ in the classic lottery model have a mechanistic basis in terms of fecundity and territorial contests. However, the drift in genotype frequencies that are the focus of the Wright-Fisher model are omitted.

@@
